Foundational claims of group polarization are plausibly false, with lessons for the empirical study of psychological dynamics

### 2024-11-09 GroupPolarizationStatmod/paper/ms.tex
added model sketches and docs to scripts

Experimetnal designs in group polarization studies share a certain
experimental boilerplate (Figure 1 --- NEED TO INCLUDE) with the goal of
testing the group polarization hypothesis that a group of like-minded
people will develop more extreme opinions on some salient topic over the
course of discussing some topic. Group polarization experiments are thus
initialized by forming like-minded groups of people and surveying their
initial opinions on some topic. In some cases these initial opinions are
used for assigning participants to like-minded groups, but other methods
have been used, such as geographic location for producing, on average,
like-minded groups \autocite{Schkade2010}. Participants then discuss the
topic in their groups. After discussion, participant opinions are again
measured. Several studies we analyze here determined group polarization
occurred (i.e., defined a positive detection) only as a non-zero
difference in mean opinion, with no other statistical testing. Some
others performed null-hypothesis \(t\)-tests to test whether there was
enough evidence to reject the null-hypothesis that \emph{simple
consensus} occurredand not group polarization. In our meta-analysis
simulations, we measure simulated group polarization effect size in
terms of Cohen's \(d\), which is an abstraction beyond these two
practices from the group polarization literature that can also be used
to compare model fits between metric models that treat group
polarization measurements as continuous although they are ordinal, and
more appropriate statistical models that account for data ordinality.

>>>>

Experimetnal designs in group polarization studies share a certain
experimental boilerplate (Figure 1 --- NEED TO INCLUDE) with the goal of
testing the group polarization hypothesis that a group of like-minded
people will develop more extreme opinions on some salient topic over the
course of discussing some topic (Figure~\ref{fig:ModelSketch}). 
Group polarization experiments are thus
initialized by forming like-minded groups of people and surveying their
initial opinions on some topic. In some cases these initial opinions are
used for assigning participants to like-minded groups, but other methods
have been used, such as geographic location for producing, on average,
like-minded groups \autocite{Schkade2010}. Participants then discuss the
topic in their groups. After discussion, participant opinions are again
measured. Several studies we analyze here determined group polarization
occurred (i.e., defined a positive detection) only as a non-zero
difference in mean opinion, with no other statistical testing. Some
others performed null-hypothesis \(t\)-tests to test whether there was
enough evidence to reject the null-hypothesis that \emph{simple
consensus} occurredand not group polarization. In our meta-analysis
simulations, we measure simulated group polarization effect size in
terms of Cohen's \(d\), which is an abstraction beyond these two
practices from the group polarization literature that can also be used
to compare model fits between metric models that treat group
polarization measurements as continuous although they are ordinal, and
more appropriate statistical models that account for data ordinality.

\begin{figure}
  \caption{\textbf{Common components of a group polarization experiment.}
  Participants first give their opinion in response to some prompt (A). Afterwards,
  participants deliberate the topic (B). Participants then give their opinion again
  (C). Opinions are implicitly behavioral choices that are most consistent with some
  complex aggregation of brain states and dynamics, and so are theoretically
  continuous (D).}
  \centering
    \includegraphics[width=0.7\textwidth]{Figures/ModelSketchesA-C.pdf}
    \includegraphics[width=0.65\textwidth]{Figures/ModelSketchesD.pdf}
  \label{fig:ModelSketch}
\end{figure}
